High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2022-01-14 PNAS_manuscript/PNAS_manuscript.tex

\subsection*{Cooling rate correction and secular variations}
Paleointensity estimates from slowly cooled igneous intrusions can also be biased toward higher values due to the orders of magnitude difference in natural and laboratory cooling rates \cite{Halgedahl1980a}. It has been shown that large differences in cooling rates between acquisition of an NRM in nature versus a pTRM in the lab can result in overestimated paleointensities for SD grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref. \citealp{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully blocked in their characteristic natural remanence magnetization ($\sim$500\textdegree C; ref. \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim 1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at $\sim 1.3\times10^{-1}$ $^\circ$C s$^{-1}$, assuming that it takes $\sim$1000s in lab condition to cool through the same temperature interval. This significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of ref. \citealp{Halgedahl1980a} (Fig. S3). This estimate on cooling rate effect is similar to the value of $\sim$35\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level weighted means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{PINT_compilation.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual (axial) dipole moment values from the PINT database (PINT v8.0.0; \url{http://www.pintdb.org/}; ref. \citealp{Bono2021a}) with additional Mesoproterozoic data from \cite{Sprain2018a, Kodama2019a}, as well as those from the Neoproterozoic \cite{Lloyd2021a, Lloyd2021b, Thallner2021a, Thallner2021b}. Paleointensity estimates from \cite{Pesonen1983a} and \cite{Kulakov2013a} are not included in the compilation due to the specimen-level double-slope behavior as discussed in the text. Overall the anorthosite xenoliths from this study record a high Mesoproterozoic field exceeding the value projected by the second order polynomial curve from \cite{Bono2019a} which is based on an interpretation of there being a progressive decay of the geodynamo through the Proterozoic.}}
\label{fig:PINT_compilation}
\end{figure*}

\section*{Proterozoic geodynamo}
To best characterize the geomagnetic axial dipole field intensity during a certain time period requires a paleointensity dataset to cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged and the geocentric axial dipole hypothesis can be assumed. Thermal modeling results from ref. \citealp{Zhang2021b} suggest that the Beaver River anorthosite xenoliths were likely heated to tholeiitic magma temperature and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a few thousand years. In addition, anorthosite xenoliths have paleointensity results that are consistent within small regions but vary between localities. Anorthosite AX12 and AX16 have very similar paleointensity estimates that are distinct from those of sites AX11 and AX13 which were emplaced $\sim$10 km away (Fig. \ref{fig:Geologic_map}). This is consistent with the interpretation that differential cooling history due to heterogeneity of the timing and depths of the Beaver River diabase intrusions caused the anorthosites to capture paleosecular variations \textit{ca.} 1092 Ma. 
% Overall, none of the anorthosite xenoliths record a site mean VDM value lower than 40 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}).

The new paleointensity records from the \textit{ca.} 1092 Ma Midcontinent Rift anorthosite xenoliths together with previous results from the \textit{ca.} 1140 Ma rocks of the Midcontinent Rift region indicate that a six-fold increase in geomagnetic dipole moment happened within $\sim$50 Myr during the late Mesoproterozoic (Fig. \ref{fig:PINT_compilation}). The Abitibi dikes \cite{Macouin2003a} record a low average virtual dipole moment of $\sim$13 ZAm$^2$, similar to the values of $\sim$7 ZAm$^2$ obtained from the single silicate crystals of the \textit{ca.} 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a}. However, the relatively high moments of $\sim$72 ZAm$^2$ from the \textit{ca.} 1105 Ma Osler Volcanic Group \cite{Sprain2018a} and $\sim$89 ZAm$^2$ from the \textit{ca.} 1092 Ma Beaver River anorthosite xenoliths (Fig. \ref{fig:PINT_compilation}) are strikingly higher than the values of the Abitibi dikes. The scale of this increase is similar to the increasing trend observed near the Ediacaran-Cambrian which has been suggested to evidence the onset of Earth's inner core nucleation \cite{Thallner2021a}.

The high geomagnetic field estimates from the Midcontinent Rift volcanics and intrusives are not the only suite of high paleointensity records at odds with the hypothesis of a monotonically decaying geodynamo through the entire Proterozoic Eon \cite{Biggin2009a, Bono2019a}. The $\sim$60 ZAm$^2$ virtual dipole moment estimates from the \textit{ca.} 755 Ma Mundine Well dikes are much higher than the values predicted by the monotonic decrease model of \cite{Bono2019a}. Although one may consider these data capturing relatively short term fluctuations of the geomagnetic field, they nevertheless indicate that the geodynamo had strong power sources during the Proterozoic. On the other hand, besides the low paleointensity values recorded by the \textit{ca.} 1140 Ma Abitibi dikes, low estimates have been obtained from the \textit{ca.} 1070 Ma Bangemall Sills \cite{Lloyd2021b}, the \textit{ca.} 720 Ma dolerite dikes of the Franklin Large Igneous Province \cite{Lloyd2021a}, as well as Devonian-aged rocks from the Minusa Basin, Kola Peninsula, Buribay volcanic complex and volcanic rocks from the East European Craton (Fig. \ref{fig:PINT_compilation}; ref. \citealp{Shcherbakova2017a, Shcherbakova2019a, Shcherbakova2021a}). The multiple occurrences of low geomagnetic field intensities during the Proterozoic and the Paleozoic indicate that a waning thermal convection regime in Earth's before the emergence of the solid inner core may not be the solely possible mechanism for triggering an weak dipole field.

In addition, a five-fold fluctuation of the geomagnetic dipole moment exist in the paleointensity records for the past 10 million years \cite{Bono2021a}, when we know a modern-day-like geodynamo likely powered by both thermal and compositional convection existed. The observation of weak fields before and after the Cambrian indicate that the geomagnetic dipole may reach particularly low strengths regardless of the existence of a solid inner core. Therefore, low paleointensity values or short-term fluctuations may not be sufficient for interpreting Earth's inner core status through history.

Recent advancements in high-pressure, high-temperature experiments for measuring thermal conductivity of iron alloys under conditions similar to Earth's core may shed new light on the age of the inner core. However, large uncertainties still remain regarding the applicability Wiedemann-Franz law of extrapolating thermal conductivity values based on electrical resistivity measurements by ref. \citealp{Ohta2016a}, and regarding the thermal modeling parameters used in the direct thermal conduction measurement experiment performed by ref. \citealp{Konopkova2016a}. Until further experiments and theory that can explain these contrasting results are developed, our knowledge on the age of Earth's inner core remain largely uncertain.

In the meantime, paleomagnetic directional anomalies may be another suite of intriguing observational records that is worth investigating. The anomalous directional fluctuations during the Ediacaran and Devonian \cite{Bono2015a, Kodama2021a, Shcherbakova2019a} perhaps can be consistent with there being a weak geodynamo regime when dipole component of Earth's field was largely diminished.

Regardless, the high values from the Midcontinent Rift intrusions and extrusive volcanics are inconsistent with the prediction of there being a progressive decay at the time (Fig. \ref{fig:PINT_compilation}; ref. \citealp{Bono2019a}) and necessitates that there being a vigorous late Mesoproterozoic geodynamo.

>>>>

\subsection*{Averaging secular variations and applying cooling rate correction}
To best characterize the geomagnetic axial dipole field intensity during a certain time period requires a paleointensity dataset to cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged and the geocentric axial dipole hypothesis can be assumed. Thermal modeling results from ref. \citealp{Zhang2021b} suggest that the Beaver River anorthosite xenoliths were likely heated to tholeiitic magma temperature and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a few thousand years. In addition, anorthosite xenoliths have paleointensity results that are consistent within small regions but vary between localities. Anorthosite AX12 and AX16 were emplaced only $\sim$450 meters apart and have very similar paleointensity estimates. $\sim$10 km to the north, anorthosite AX11 and AX13 were also emplaced closely at a distance of $\sim$125 meters but yielding lower paleointensity estimates from those of AX12 and AX16 (Fig. \ref{fig:Geologic_map}). This is consistent with the interpretation that differential cooling history due to heterogeneity of the timing and depths of the Beaver River diabase intrusions caused the anorthosites to capture paleosecular variations \textit{ca.} 1092 Ma.

Another consequence of having the diabase and anorthosite cooled slowly during emplacement is its effect on biasing the paleointensity estimates toward higher values. It has been shown that large differences in cooling rates between acquisition of an NRM in nature versus a pTRM in the lab can result in overestimated paleointensities for SD grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref. \citealp{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully blocked in their characteristic natural remanence magnetization ($\sim$500\textdegree C; ref. \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim 1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at $\sim 1.3\times10^{-1}$ $^\circ$C s$^{-1}$, assuming that it takes $\sim$1000s in lab condition to cool through the same temperature interval. This significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of ref. \citealp{Halgedahl1980a} (Fig. S3). This estimate on cooling rate effect is similar to the value of $\sim$35\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level weighted means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{PINT_compilation.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual (axial) dipole moment values from the PINT database (PINT v8.0.0; \url{http://www.pintdb.org/}; ref. \citealp{Bono2021a}) with additional Neoproterozoic data from refs. \citealp{Lloyd2021a, Lloyd2021b, Thallner2021a, Thallner2021b}. Paleointensity estimates from \cite{Pesonen1983a} and \cite{Kulakov2013a} are not included in the compilation due to the specimen-level double-slope behavior as discussed in the text. Overall the anorthosite xenoliths from this study record a high Mesoproterozoic field exceeding the value projected by the second order polynomial curve from \cite{Bono2019a} which is based on an interpretation of there being a monotonic decay of the geodynamo through the Proterozoic.}}
\label{fig:PINT_compilation}
\end{figure*}

\section*{Proterozoic geodynamo}

The long-term evolution of the core's thermal regime and geodynamo has been an outstanding question in the study of Earth history. Progress has been made in several fronts including constraining iron and its alloy's thermal conductivity values in high-temperature, high-pressure conditions similar to that in Earth's core; constructing numerical geodynamo models to simulate geomagnetic field behaviors under various thermal regimes; and obtaining and compiling observational paleomagnetic intensity records to test hypotheses put forward by numerical models.

Some studies interpret that the records of the Proterozoic paleointensities hint a progressive, monotonic decay trend, indicating that the geodynamo had a progressively waning power source \cite[e.g.]{Bono2019a}. This trend was interpreted to support the hypothesis that Earth's inner core had not solidified and thermal convection was dissipating away its rigor during the Proterozoic. With ultra-low paleointensity observations obtained from the ca. 565 Ma Sept-$\hat{I}$les intrusions, ref. \citealp{Bono2019a} hypothesized that the inner core did not begin to nucleate before at least the late Neoproterozoic and the thermal conductivity of the core must be near the high end of viable values obtained through experiments \cite{Ohta2016a}. Following this hypothesis, numerical modeling with particularly high core thermal conductivity values show that a decaying geodynamo through the Proterozoic would reach its lowest strength before the emergence of the solid inner core during the Ediacaran-Cambrian \cite{Davies2021a}. Modeling also predicts that a rapid (within a few 10s of Myr) recovery in geomagnetic field intensity occurred soon after the onset of inner core nucleation, as compositional convection started to provide additional power to drive the geodynamo \cite{Driscoll2016a}.

However, the high-quality, consistent paleointensity estimates from the well-preserved ca. 1092 Ma anorthosite xenoliths of the intracontinental Midcontinent Rift confirm the existence of a high geomagnetic field in the late Mesoproterozoic. Together with previous records obtained from the ca. 1106 Ma Osler Volcanics of the Midcontinent Rift \cite{Sprain2018a}, these data indicate an appreciable power source for Earth's dynamo could have persisted through at least 14 Myr during the Mesoproterozoic (Fig. \ref{fig:PINT_compilation}). Besides the high geomagnetic fields recorded by the Midcontinent Rift rocks, the ca. 755 Ma Mundine well dikes also record a strong geomagnetic field in the Neoproterozoic with an average virtual dipole field of $\sim$60 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}). These observations are inconsistent with there being a monotonic decay during the Proterozoic as predicted by magnetic scaling and thermal evolution models using high core thermal conductivity values \cite{Aubert2009a, Davies2021a}.

Numerical simulations that apply magnetic scaling law also predict that there was a strong increase in true geomagnetic dipole moment during the early stages of inner core growth \cite{Davies2021a}. The updated paleointensity data compilation shows that rapid transitions from low to high field intensities do not occur on an unique case during Earth history (Fig. \ref{fig:PINT_compilation}). A two-fold increase in Earth's virtual dipole moment in the Ediacaran has been recorded by the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} and the ca. 550 Ma volcanics rocks of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). That transition has been interpreted to be caused by the initiation of the nucleation of the solid inner core \cite{Thallner2021a} . However, in the Mesoproterozoic, a six-fold increase in geomagnetic dipole moment is observed within a short period ($\sim$40 Myr) from a very low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, which might have lasted through ca. 1092 Ma when the Beaver River anorthosite xenoliths recorded a virtual dipole moment of $\sim$80 ZAm$^2$. This increase in absolute value of virtual dipole moment dwarfs all other existing records of geomagnetic field fluctuations in the Precambrian, including those in the Ediacaran (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition may also be a better match with that predicted by numerical simulations upon the onset of inner core nucleation \cite{Davies2021a} but this similarity could be coincidental.

On the other hand, particularly low paleointensity values similar to those observed in the Ediacaran has been obtained at multiple times before and after the Ediacaran-Cambrian. In the late Mesoproterozoic, besides the the \textit{ca.} low intensities recorded by the ca. 1140 Ma Abitibi dikes \cite{Macouin2003a}, one ca. 1070 Ma Bangemall Sill records an extremely low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; ref. \citealp{Lloyd2021b}). In the Neoproterozoic, ca. 720 Ma dolerite dikes of the Franklin Large Igneous Province record a mean virtual dipole moment of 11 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; ref. \citealp{Lloyd2021a}). In the Phanerazoic, Devonian-aged rocks from the Strathmore lava flows \cite{Hawkins2021a}, Minusa Basin, Kola Peninsula \cite{Shcherbakova2017a}, volcanic rocks from the East European Craton \cite{Shcherbakova2019a}, as well as the Buribay volcanic complex \cite{Shcherbakova2021a} show low paleointensities through an extended period of time (Fig. \ref{fig:PINT_compilation}). Accompanying the low paleointensity values, paleodirectional data from the Ediacaran-Cambrian \cite{Bono2015a, Halls2015b, Kodama2021a} and the Devonian show anomalous directional fluctuations \cite{Shcherbakova2019a} which could be consistent with there being a largely diminished dipole component of Earth's magnetic field. If the weak paleointensities in the Ediacaran or the Devonian are attributed to the geodynamo being on the brink of collapse just prior to the nucleation of the solid inner core, such data would suggest that the thermal conductivity of the core must be extremely high \cite{Davies2015a, Ohta2016a}. Alternatively, these Proterozoic and Paleozoic records could be a manifestation of cyclic forcing of the geodynamo across the core–mantle boundary with a time scale on the order of 150 to 200 Myr \cite{Courtillot1987a, Hawkins2021a}, a feature that has been proposed to be associated with the Wilson cycle and magnetic superchrons, and perhaps postdate the emergence of the solid inner core \cite{Courtillot1987a, Courtillot2007a}.

Another alternative explanation for the low paleointensities in the Paleozoic and Proterozoic is that they could be a result of short-term geomagnetic field fluctuations. A five-fold fluctuation of the geomagnetic dipole moment exist in the paleointensity records for the past 10 million years \cite{Bono2021a}, when we know a modern-day-like geodynamo likely powered by both thermal and compositional convection existed. The observation of weak fields before and after the Cambrian indicate that the geomagnetic field may excurse toward particularly low strengths on narrow time scales regardless of the existence of a solid inner core. Therefore, low paleointensity values or short-term fluctuations may not be sufficient for interpreting Earth's inner core status through history.

% The high geomagnetic field estimates from the Midcontinent Rift volcanics and intrusives are not the only suite of high paleointensity records at odds with the hypothesis of a monotonically decaying geodynamo through the entire Proterozoic Eon \cite{Biggin2009a, Bono2019a}. The $\sim$60 ZAm$^2$ virtual dipole moment estimates from the \textit{ca.} 755 Ma Mundine Well dikes are much higher than the values predicted by the monotonic decrease model of \cite{Bono2019a}. Although one may consider these data capturing relatively short term fluctuations of the geomagnetic field, they nevertheless indicate that the geodynamo had strong power sources during the Proterozoic. On the other hand, besides the low paleointensity values recorded by the \textit{ca.} 1140 Ma Abitibi dikes, low estimates have been obtained from the \textit{ca.} 1070 Ma Bangemall Sills \cite{Lloyd2021b}, the \textit{ca.} 720 Ma dolerite dikes of the Franklin Large Igneous Province \cite{Lloyd2021a}, as well as Devonian-aged rocks from the Minusa Basin, Kola Peninsula, Buribay volcanic complex and volcanic rocks from the East European Craton (Fig. \ref{fig:PINT_compilation}; ref. \citealp{Shcherbakova2017a, Shcherbakova2019a, Shcherbakova2021a}). The multiple occurrences of low geomagnetic field intensities during the Proterozoic and the Paleozoic indicate that a waning thermal convection regime in Earth's before the emergence of the solid inner core may not be the solely possible mechanism for triggering an weak dipole field.

Constraining the age of the inner core remains to be an open question given the uncertainties in measuring the thermal conductivity values of iron alloys in lab using different approaches \cite{Konopkova2016a, Ohta2016a}. Experimental challenges such as measuring sample thicknesses under high-pressure, high-temperature conditions, together with theoretical difficulties of the validity of applying the Wiedemann-Franz law to extrapolate thermal conductivity values based on electrical resistivity measurements \cite{Ohta2016a}, and modeling uncertainties associated with parameters used in the direct thermal conduction measurement experiment \citea{Konopkova2016a} cast uncertainty in the age of the inner core by hundreds of millions of years. Until further experiments and theory that can explain these contrasting results are developed, our knowledge on the age of Earth's thermal evolution remain largely uncertain.

% In the meantime, paleomagnetic directional anomalies may be another suite of intriguing observational records that is worth investigating. The anomalous directional fluctuations during the Ediacaran and Devonian \cite{Bono2015a, Kodama2021a, Shcherbakova2019a} perhaps can be consistent with there being a weak geodynamo regime when dipole component of Earth's field was largely diminished.

Overall, the high-fidelity paleointensity recorders of the Beaver River anorthosite xenoliths in the well-preserved Midcontinent Rift record high paleointensities values and necessitates that there was a vigorous late Mesoproterozoic geodynamo.
